Constraint-Preserving IMEX Scheme for the Two-Moment Model of Fermion Transport

### 2018-09-19 Documents/M1/Astronum_2018/sections/Numerical.tex
Commited for backup.

\subsection{Accuracy Tests}

@@

>>>>

\subsection{Accuracy Tests}
To compare the accuracy of the IMEX scheme, we applied our PDARSs with RKCB2 [citation] and SSP2332[citation] scheme to some smooth problems in streaming, absorption, ans scattering dominated regimes in one spatial dimension.
All the tests in this subsection were applied with third order accurate spatial discretization, the maximum entropy closure in the low occupancy limit and time step $\dt = 0.1 \times \dx $.
More information and the definition of the absolute error and the relative error can be found in \cite{Chu_2018}.

\subsubsection{Sine Wave Streaming}
This test involves the streaming part only: the collision term is turned off. 
A periodic domain $D$ was applied and the initial condition 
\begin{figure}[h]
  \centering
    \includegraphics[width=0.6\textwidth]{figures/SineWaveStreaming}
   \caption{Sine Wave Streaming}
\end{figure}

\subsubsection{Sine Wave Damping}
\begin{figure}[h]
  \centering
    \includegraphics[width=0.7\textwidth]{figures/SineWaveDamping}
   \caption{Sine Wave Damping}
\end{figure}
\subsubsection{Sine Wave Diffusion}
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/SineWaveDiffusionJ}
    \includegraphics[width=0.5\textwidth]{figures/SineWaveDiffusionH}
  \end{tabular}
   \caption{Sine Wave Diffusion}
\end{figure}

@@

\subsubsection{Neutrino Stationary State Test}
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.4\textwidth]{figures/NStatinaryS_EOS}
    \includegraphics[width=0.55\textwidth]{figures/NSS_Opacities}
  \end{tabular}
   \caption{Neutrino Stationary State Test}
\end{figure}

\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.45\textwidth]{figures/NSS_0_1}
    \includegraphics[width=0.45\textwidth]{figures/NSS_0_2} \\
    \includegraphics[width=0.45\textwidth]{figures/NSS_1_1}
    \includegraphics[width=0.45\textwidth]{figures/NSS_1_2} \\
    \includegraphics[width=0.45\textwidth]{figures/NSS_3_1}
    \includegraphics[width=0.45\textwidth]{figures/NSS_3_2} \\
    \includegraphics[width=0.45\textwidth]{figures/NSS_5_1}
    \includegraphics[width=0.45\textwidth]{figures/NSS_5_2} \\
  \end{tabular}
   \caption{Neutrino Stationary State Test}
\end{figure}

### 2018-09-19 Documents/M1/Astronum_2018/sections/SpacialDiscretization.tex
Commited for backup.

\section{Spacial Discretization}
Let the spatial domain be $D$ and divide it into $N$ cells, denoted $I_{i}$ with $i = 1,\ldots,N$:
\begin{equation*}
D = \cup_{i = 1}^{N} I_{i}
\end{equation*}
\begin{equation}
\bcM_{i} = \dfrac{1}{\dx} \int_{I_i}\bcM dx
\end{equation}

\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) +  \cC(\bcM_{i})
\end{equation}

\begin{equation}
  \widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big)
\end{equation}

By treating the flux term explicitly and the collision term implicitly, we have
\begin{align}
\bcM_{i}^{n+1} 
& = \bcM_{i}^{n} - \dfrac{\dt}{\dx} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right) + \dt\,\cC(\bcM^{n+1}_{i}) \nonumber \\
& = (1-\eta)\bcM_{i}^{n} + \f{1}{2}\eta\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\eta\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right) + \dt\,\cC(\bcM^{n+1}_{i})
\end{align}
with $\eta = \dfrac{\dt}{\dx}$. The first three terms can form a convex combination.

>>>>

\section{Spacial Discretization}
Divide the spatial domain $D$ into $N$ cells and denote the cell as $I_{i}$ with $i = 1,\ldots,N$, so that
\begin{equation*}
D = \cup_{i = 1}^{N} I_{i}.
\end{equation*}
Each of the cell has a volume/length of $\dx$.
Then we have cell average moment:
\begin{equation}
\bcM_{i} = \dfrac{1}{\dx} \int_{I_i}\bcM dx.
\end{equation}
Integrate Eq.~\eqref{eq:momentEquations} for each cell $I_{i}$ to have
\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the integrated collision term.
In this paper we use the simple Lax-Friedrichs flux that given by
\begin{equation}
  \widehat{\bcF}_{LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
\end{equation}
Therefore, by treating the flux term explicitly and the collision term implicitly, we have
\begin{align}
\bcM_{i}^{n+1} 
& = \bcM_{i}^{n} - \dfrac{\dt}{\dx} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i}) \nonumber \\
& = (1-\eta)\bcM_{i}^{n} + \f{1}{2}\eta\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\eta\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i})
\end{align}
where $\eta := \dfrac{\dt}{\dx}$. The first three terms can form a 3-term convex combination by requiring $\eta \in [0,1]$.
It can be shown that $\bcM_{i}^{n+1}$ remains realizable given $\bcM_{i}^{n}$ is realizable, $\eta \in [0,1]$ and $\cC(\bcM)$ is positivity-preserving. The proof can be found in \cite{Chu_2018}.

### 2018-09-19 Documents/M1/Astronum_2018/sections/TimeIntegration.tex
Commited for backup.

\section{Time Integration}

Let $\vect{\cQ}$ be some linear operators and
\begin{equation}
\vect{u}(t) = \left( \bcM_{1}(t),\ldots,\bcM_{N}(t)\right) ^{T}
\end{equation}
\begin{equation}
  \dfrac{d \vect{u}}{d t} = \vect{\cT}(\vect{u}) + \vect{\cQ}(\vect{u})
\end{equation}
\subsection{Standard Implicit-Explicit Scheme}

\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \tilde{\vect{u}}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} 
\end{align}

\subsection{Constraint-Preserving Implicit-Explicit Scheme}

\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}

\subsubsection{PD-ARS with SSPRK2}

\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2-\epsilon & 1/2+\epsilon \\ \hline
  	  & 0 & 1/2-\epsilon & 1/2+\epsilon
  \end{array}
\end{equation}

@@

\subsubsection{PD-ARS with SSPRK3}
\begin{equation}
  \begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6-\epsilon^2 /4 & 1/6-\epsilon/4 & 2/3+\epsilon(1+\epsilon)/4\\\hline
  	  & 0 & 1/6-\epsilon^2 /4 & 1/6-\epsilon/4 & 2/3+\epsilon(1+\epsilon)/4
  \end{array}
\end{equation}
For these schemes, $c_{\mbox{\tiny Sch}}= 1 - 3\epsilon/2$ with $\epsilon \rightarrow 0^{+}$.

>>>>

\section{Time Integration}
In this section we show how to use the property we introduced at the end of the previous section to construct a constraint-preserving implicit-explicit (IMEX) scheme.
We can have a more general form of the partial differential equation, Eq.~\eqref{eq:momentEquations},
\begin{equation}
  \dfrac{d \vect{u}}{d t} = \vect{\cT}(\vect{u}) + \f{1}{\tau}\,\vect{\cQ}(\vect{u}),
\end{equation}
where
\begin{equation}
\vect{u}(t) = \left( \bcM_{1}(t),\ldots,\bcM_{N}(t)\right) ^{T}
\end{equation}
is the collection of all moment cell average, $\vect{\cT}$ is the operator corresponding to the flux term and $\f{1}{\tau}\,\vect{\cQ}$ is a linear operators corresponding to the collision term.

\subsection{Standard Implicit-Explicit Scheme}
Treating $\vect{\cT}$ explicitly and $\f{1}{\tau}\,\vect{\cQ}$ implicitly, the standard $s$-stage IMEX schemes take the following form 
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \vect{u}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} 
\end{align}
where  $(\tilde{a}_{ij})$ and $(a_{ij})$ are coefficients of the $i$th-stage and they are the elements of matrices $\tilde{A}$ and $A$, respectively, and the vectors $\tilde{\vect{w}}=(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $\vect{w}=(w_{1},\ldots,w_{s})^{T}$ are the weights vector.
Some conditions those coefficients and weights have to satisfy to serve an accurate, converge and stable scheme.
For instance, to have second-order temporal accuracy, the following conditions are required:
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=\sum_{i=1}^{s}w_{i}=1,
  \label{orderConditions1}
\end{equation}
and
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}\tilde{w}_{i}\,c_{i}
  =\sum_{i=1}^{s}w_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}w_{i}\,c_{i}=\f{1}{2}, 
  \label{orderConditions2}
\end{equation}
where $\tilde{c}_{i} = \sum_{j}^{s}\tilde{a}_{ij}$ and $c_{i}=\sum_{j}^{s}a_{ij}$.

\subsection{Constraint-Preserving Implicit-Explicit Scheme}
The problem left for constructing a constraint-preserving IMEX is finding the conditions $(\tilde{a}_{ij})$, $(a_{ij})$, $\tilde{w}_{i}$ and $w_{i}$ need to satisfied and giving their possible values.
As Hu et al. did in their paper \cite{hu_etal_2018}, the stage values in Eq.~\eqref{imexStages} can be rewritten to
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
It can be proved \cite{Chu_2018} that the conditions we need for constraint-preserving IMEX is:
\begin{enumerate}
    \item Consistency of the implicit coefficients
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item Second-order accuracy in the streaming limit; i.e., satisfies \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
      \label{eq:orderConditionsEx}
    \end{equation}
    \item Convex-invariant; i.e. satisfies 
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Well-behaved in the diffusion limit; i.e., satisfies \begin{equation}
      \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Less than four stages ($s\le3$).
    \item Globally stiffly accurate (GSA): $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$.
\end{enumerate}  
We call the IMEX scheme satisfying the above conditions {PD-ARS}. (Definition 3 in \cite{Chu_2018})
We give two optimized PD-ARS scheme here: PD-ARS with SSPRK2 and  PD-ARS with SSPRK3. They have $2$-nd order and $3$-rd order accuracy in the streaming limit, respectively.
\subsubsection{PD-ARS with SSPRK2}
The explicit part of this IMEX scheme is SSPRK2.
\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2-\epsilon & 1/2+\epsilon \\ \hline
  	  & 0 & 1/2-\epsilon & 1/2+\epsilon
  \end{array}
\end{equation}

@@

\subsubsection{PD-ARS with SSPRK3}
The explicit part of this IMEX scheme is SSPRK3.

\begin{equation}
  \begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6-\epsilon^2 /4 & 1/6-\epsilon/4 & 2/3+\epsilon(1+\epsilon)/4\\\hline
  	  & 0 & 1/6-\epsilon^2 /4 & 1/6-\epsilon/4 & 2/3+\epsilon(1+\epsilon)/4
  \end{array}
\end{equation}
For these schemes, $c_{\mbox{\tiny Sch}}= 1 - 3\epsilon/2$ with $\epsilon \rightarrow 0^{+}$.

### 2018-09-19 Documents/M1/Astronum_2018/sections/TwoMoments.tex
Commited for backup.

\subsection{Constraints on Moments}
One fundamental fact one should concern before solving the system is the property he/she should never violate.
For a fermionic system, the property is Pauli exclusion principle, or 
\begin{equation}
f \in (0,1).
\end{equation}
As an result, the angular moments as the integration of $f$ are confined as following
\begin{align}
\cJ \in(0,1), \quad &(1-\cJ)\cJ-|\vect{\cH}|  > 0, \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  < & \chi<\min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}}.  
  \label{eq:eddingtonFactorBounds}
\end{align}
As Fig.~\eqref{fig:EddingtonFactorsWithDifferentClosure} shows, not all the algebraic closures satisfy the constraints on moments all the time: Kershaw [citation], Wilson [citation], Levermore [citation], Minerbo [citation], Janka 2 [citation] may work fine at low occupancy case, but not at high occupancy situation; Janka 1 [citation] is even worse and not work all the time. Only Cernohorsky \& Bludman[citation] among these plotted closures remains in the realizable region.
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_10}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_90}
  \end{tabular}
   \caption{Plot of Eddington factors $\chi$ versus flux factor $h$ for different values of $\cJ$ for various algebraic closures: $\cJ=0.1$ (left panel, low occupied) and $\cJ=0.9$ (right panel, high occupied).  In each panel we plot the Eddington factors of Kershaw (red), Wilson (yellow), Levermore (green), Minerbo (light blue), Cernohorsky \& Bludman (blue) and Janka (purple and pink) closures.  We also plot $\chi_{\mbox{\tiny min}}$ and $\chi_{\mbox{\tiny max}}$ defined in Eq.~\eqref{eq:eddingtonFactorBounds} (lower and upper dash black lines, respectively).}
  \label{fig:EddingtonFactorsWithDifferentClosure}
\end{figure}

>>>>

\subsection{Constraints on Moments}
One fundamental fact one should concern before solving the system is the property he/she should never violate.
For a fermionic system, the property is Pauli exclusion principle, or 
\begin{equation}
f \in (0,1).
\end{equation}
As a result, the angular moments as the integrations of $f$ are confined as following
\begin{align}
\cJ \in(0,1), \quad &(1-\cJ)\cJ-|\vect{\cH}|  > 0, \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  < & \chi<\min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.
As Fig.~\eqref{fig:EddingtonFactorsWithDifferentClosure} shows, not all the algebraic closures satisfy the constraints on moments all the time: Kershaw [citation], Wilson [citation], Levermore [citation], Minerbo [citation], Janka 2 [citation] may work fine at low occupancy case, but not at high occupancy situation; Janka 1 [citation] is even worse and not work all the time. Only Cernohorsky \& Bludman[citation] among these plotted closures remains in the realizable region.
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_10}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_90}
  \end{tabular}
   \caption{Plot of Eddington factors $\chi$ versus flux factor $h$ for different values of $\cJ$ for various algebraic closures: $\cJ=0.1$ (left panel, low occupied) and $\cJ=0.9$ (right panel, high occupied).  In each panel we plot the Eddington factors of Kershaw (red), Wilson (yellow), Levermore (green), Minerbo (light blue), Cernohorsky \& Bludman (blue) and Janka (purple and pink) closures.  We also plot $\chi_{\mbox{\tiny min}}$ and $\chi_{\mbox{\tiny max}}$ defined in Eq.~\eqref{eq:eddingtonFactorBounds} (lower and upper dash black lines, respectively).}
  \label{fig:EddingtonFactorsWithDifferentClosure}
\end{figure}
